Well-quasi-orderings on word languages

### 2024-12-16 src/infixes.tex
New theorems

\begin{conjecture}
    \label{small-ordinal-invariants:conj}
    Let $L$ be a \kl{well-quasi-ordered} language for the \kl{infix relation}
    that is \kl{downwards closed}.
    Then, the \kl{maximal order type} of $L$ is at most $\omega^3$,
    its \kl{ordinal height} is at most $\omega$,
    and its \kl{ordinal width} is at most $\omega^2$.
\end{conjecture}

>>>>

\begin{conjecture}
    \label{small-ordinal-invariants:conj}
    Let $L$ be a \kl{well-quasi-ordered} language for the \kl{infix relation}
    that is \kl{downwards closed}.
    Then, the \kl{maximal order type} of $L$ is at most $\omegaOrd^3$,
    its \kl{ordinal height} is at most $\omegaOrd$,
    and its \kl{ordinal width} is at most $\omegaOrd^2$.
\end{conjecture}

\todo[inline]{
    Write this in a nicer way.
    Maybe talk a bit more about automatic sequences?
    This could be a section ``Infixes and downwards closed languages''. Before the
    ``infixes and amalgamation systems''.
}

\begin{theorem}
    \label{small-ordinal-invariants:thm}
    Let $L$ be a \kl{well-quasi-ordered} language for the \kl{infix relation}
    that is \kl{downwards closed}.
    Then, the \kl{maximal order type} of $L$ is strictly less than $\omegaOrd^3$,
    its \kl{ordinal height} is at most $\omegaOrd$,
    and its \kl{ordinal width} is at most $\omegaOrd^2$.
\end{theorem}

The \cref{small-ordinal-invariants:thm} is a direct consequence of the following
remarks.

\begin{itemize}
    \item If $L$ is \kl{wqo} and \kl{downwards closed}, then $L = \bigcup_{i = 1}^n L_i$ where $L_i$
        is an \kl{order ideal}.
    \item If $L$ is an \kl{order ideal} that is \kl{well-quasi-ordered}, then
        $L$ is the set of infixes of some \emph{bi-infinite} word $w \in \Sigma^{\Rel}$.
    \item The set of infixes of a \emph{bi-infinte} word can be understood 
        as the set of \kl{infixes} of two infinite words $u,v \in \Sigma^{\Nat}$.
    \item The set of infixes of an infinite word $w \in \Sigma^{\Nat}$ is 
        \kl{well-quasi-ordered} if and only if $w$ is \kl{ultimately uniformly recurrent},
        that is, if there exists an infinite suffix of $w$ that is \kl{uniformly recurrent}.
    \item The ordinal invariants of a \kl{uniformly recurrent} word are small.
\end{itemize}

Note that being \kl{uniformly recurrent} still leaves open a lot of potential behaviours,
for instance, Sturmian words are \kl{uniformly recurrent}, but are not \kl{automatic}.
\todo{Find citations for this.}

\begin{theorem}
    Given an \kl{automatic sequence} $w \in \Sigma^{\Nat}$, one can decide
    whether the set of \kl{infixes} of $w$ is \kl{well-quasi-ordered} for the
    \kl{infix relation}, and if so, compute the associated \kl{ordinal
    invariants}.
\end{theorem}
\begin{proof}
    One can check if $w$ is \kl{ultimately uniformly recurrent} by checking if
    $\exists N_0 \in \Nat, \dots$. Because $w$ is \kl{automatic}, this is decidable.
    Now, to decide whether $w$ has width $\omegaOrd$, then one can check if $w$ is 
    \kl{ultimately periodic}. If not, then $w$ has width $\omegaOrd$.
    If it is \kl{ultimately periodic}, one can compute the smallest size of
    the period, which gives us the width.
    Using the width, and because we know the height is at most $\omegaOrd$, we can
    compute the maximal order type.
\end{proof}

### 2024-12-16 src/introduction.tex
New theorems

This result directly translates into upper bounds on the possible \kl{ordinal invariants}
of such languages similarly as for the \kl{prefix relation}. Notice that these
upper bounds are significantly smaller than \kl{ordinal invariants} of the \kl{subword
relation}.
{
\renewcommand{\labelenumi}{R\arabic{enumi}}
\begin{enumerate}
    \setcounter{enumi}{4}
    \item All \kl{bounded languages}
            $L \subseteq \Sigma^*$ that are \kl{well-quasi-ordered} by the \kl{infix relation}
            have an \kl{ordinal height} of at most $\omega$,
            an \kl{ordinal width} strictly below $\omega^2$, and
            a \kl{maximal order type} strictly below $\omega^3$.
\end{enumerate}
}

>>>>

This result directly translates into upper bounds on the possible \kl{ordinal invariants}
of such languages similarly as for the \kl{prefix relation}. Notice that these
upper bounds are significantly smaller than \kl{ordinal invariants} of the \kl{subword
relation}.
\todo[inline]{Rephrase: all downwards closed languages have understandable invariants.
    And bounded languages are WQO if and only if their downward clsosure is!}
{
\renewcommand{\labelenumi}{R\arabic{enumi}}
\begin{enumerate}
    \setcounter{enumi}{4}
    \item All \kl{bounded languages}
            $L \subseteq \Sigma^*$ that are \kl{well-quasi-ordered} by the \kl{infix relation}
            have an \kl{ordinal height} of at most $\omegaOrd$,
            an \kl{ordinal width} strictly below $\omegaOrd^2$, and
            a \kl{maximal order type} strictly below $\omegaOrd^3$.
            \todo{Add reference to the corollary}
    \item All \kl{downwards closed} languages $L$
            that are \kl{well-quasi-ordered} by the \kl{infix relation}
            have an \kl{ordinal height} of at most $\omegaOrd$,
            an \kl{ordinal width} strictly below $\omegaOrd^2$, and
            a \kl{maximal order type} strictly below $\omegaOrd^3$.
            \todo{Add reference to the corollary}
\end{enumerate}
}
